Automatic Parallelization and Transparent Fault Tolerance\\ (\emph{New project} paper)

### 2016-04-22 pubs/tfp16/paper.tex
interim or final paper submission, deadline is Monday

\renewcommand{\floatpagefraction}{0.8}
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Automatic Parallelization and Transparent Fault Tolerance\\
       (\emph{New project} paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and
        Dean Prichard\inst{1} \and
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{Ringo, Anderson, and Marks are undergraduate students.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

A claimed trend is an increasing appreciation and practice of functional
programming in scientific computing, and in particular strict (by default)
functional programming, in various guises.  Besides making programs easier to
reason about, this enables automatic parallelization to various degrees.  A
second claimed trend in both scientific computing and functional programming
is transparent fault tolerance.  Our project goal is the demonstration of a
light-weight, higher-order, polymorphic, pure functional language
implementation in which we can experiment with automatic parallelization
strategies, varying degrees of strictness, and mechanisms for transparent
fault tolerance.
%
We do not consider speculative evaluation, or semantic strictness inferred by
program analysis, so potential parallelism is dictated by the degree of
strictness.\\
{\color{OliveGreen} % \bf
\emph{Draft paper notes:}
\begin{enumerate}[before=\itshape]
\item We are reporting work in progress.  At the time of writing we have a
  serial implementation, a primitive proof-of-concept parallel
  implementation, and are currently developing a more realistic parallel
  runtime.  We expect to have preliminary results from our new parallel
  implementation by the time of the symposium.  The fault-tolerance mechanisms
  remain future work.

\item Hopefully in the spirit of a student-friendly symposium, this draft is
  liberally annotated with indicators of student contributions, student
  preparation, the influence of their involvement on our approach, and other
  incidental and ancillary comments.  A final paper would of course be cleaned
  up.

\item While \textcolor{OrangeRed}{this does not qualify as a student paper}, students
  (undergraduate!) have contributed considerably to this effort, and to the
  writing of this paper, and will continue to do so.  If this paper is
  accepted we plan to have one of these students present it at TiFP'16.
\end{enumerate}
}

\keywords{Strict pure functional programming, automatic parallelization,
  transparent fault tolerance, Haskell, STG, GHC, Core.}

@@

\section{Background and Motivation}
The practice of so-called high-performance scientific computing (HPC) is,
overall, highly conservative with respect to change.  Even as the USA and
other governments commit to a push for exascale computing ($10^{18}$ FLOPS in
a single system) within the next
decade~\cite{Exascale-roadmap,EUexascale,Exascale-org} a highly respected
member of the HPC community recently stated publicly that ``Fortran is
essential for exascale programming''~\cite{Heroux16}.  Existing US government
laboratory HPC codes run to hundreds of thousands of lines of Fortran each and
complete rewrites are infeasible.\footnote{The codes of our lab's British
  counterpart are similar.}  C++ has become more common for new code starts,
but their nature remains the same: huge codes with fair to non-existent
encapsulation of side effects, with multiple levels of parallelism sometimes
poorly abstracted, especially at the thread (shared-memory) and computational
accelerator (e.g.\ GPU) levels.  Thread-level parallel efficiency can be good
when the parallel model is bulk synchronous; tasking models are generally
avoided unless abstracted by, e.g., a C++ template or runtime library, because
of the programming complexity.

@@

A common complaint among these aspiring high-performance functional
programmers is the need to get strictness just right: to remove space leaks,
achieve acceptable performance, get I/O and inter-process communication to work
as they expect, and so on.  More specifically, they expect to be able to
reason about space and time usage, and order of evaluation, using their
existing mental models. While they are happy to write code that is strongly
reminiscent of the numerical examples in Hughes' classic piece (whether
they've read it or not)~\cite{Hughes89}, they would also be happy to write in
a slightly modified style that does not require non-strictness---after all,
that's how they've been thinking their entire careers.

@@

Our initial requirements are that
%
\begin{itemize}
\item The implementation must be feasible in terms of available effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling must be properly and consistently implemented;
\item The generated code is C.
\end{itemize}
%
For feasibility we restrict ourselves to implicit parallelism on shared-memory
multiprocessors.  The requirement for tail calling is not primarily for
efficiency, rather for an efficient parallel implementation.  Achieving
this, and thereby sufficiently limiting the use of the native C stack, will be
discussed at some length.

@@

\section{Design and Implementation}
Our wish list has much in common with the GHC Haskell implementation, and
indeed GHC is part of our master plan.  In brief, the phases of the GHC
compiler are (1) parsing and type checking; (2) transformation to the GHC
\emph{Core} intermediate representation~\cite{Sulzmann:2007,ghc-core}; (3)
analysis and transformation on Core; (4) transformation of Core to the
\emph{STG} intermediate representation~\cite{PJ-stockhardware}; (5)
transformation of STG to machine code via various imperative
representations (Cmm, LLVM, C, native code generation).

@@

Greater strictness implies more opportunities for automatic parallelism,
possible losses or gains in space and time efficiency, and less expressiveness.  We
seek to explore the interplay of all of these.

GHC 8.0 gives three defaults: conventional non-strict, constructor strict, and
function- and constructor strict, but there are possible variations.  The
obvious variation is function, but not constructor, strictness.  In Haskell
the distinction between constructors and functions is somewhat blurred
%because constructors behave like functions 
(outside of data definitions and pattern matching), but in STG the distinction
is clear.

@@

In STG the arity of a function is defined as the number of formal parameters
in the function definition, thus \texttt{id x = x} has by definition arity one
regardless of the type of the argument (e.g., some function type) it might be
applied to.  We distinguish \emph{application}, e.g., \texttt{f x y}, which is
a syntactic notion, from actual function \emph{call}. Thus \texttt{id id x} is
an application of \texttt{id} to two arguments, but the underlying \emph{FUN}
(user- or system- defined function) when called is given exactly one
argument (the first one).

In the eval-apply model, given application \texttt{f x y}, in the standard
(non-strict) semantics \texttt{f} is first evaluated and the arity of the
underlying \emph{FUN} or \emph{PAP} (partial application of a \emph{FUN}) is
determined---in the case of \emph{PAP} the arity is that of the underlying
\emph{FUN} less the number of arguments previously provided.  If the arity is
greater than two a new \emph{PAP} is created with arguments \texttt{x} and
\texttt{y} inserted and the new \emph{PAP} returned.  If the arity is equal to
two the underlying \emph{FUN} is tail-called (jumped to) with any previous
arguments (in case of \emph{PAP}) and new arguments \texttt{x} and \texttt{y}.
If the arity is less than two (it must be at least one, so one in this case)
the underlying \emph{FUN} is called with any previous arguments (in case of
\emph{PAP}) and new argument \texttt{x}, then the object returned applied to
remaining argument \texttt{y}.

Given application \texttt{f x y}, when might \texttt{x} and \texttt{y} be
evaluated?  In the non-strict semantics evaluation is on demand, after
\texttt{f} has been evaluated, the underlying \emph{FUN} actually called
(i.e., not just partially applied and wrapped up in a \emph{PAP}), and one or
both of those arguments possibly forced in subsequent evaluation (assuming
that they weren't forced in the evaluation of \texttt{f} itself).

In a strict semantics one could devise various evaluation orders, but as far
as distinguishing termination properties it seems that these can be distilled
down to two reasonable possibilities, namely whether on construction of a
\emph{PAP} the (new) arguments are forced or not.  In other words, the
question is whether application is strict, or function call is strict.  These
are in fact distinguishable because (in Haskell, for example) the first
argument to \texttt{seq} could evaluate to a \emph{PAP}.

This then gives three reasonable operational modes which are strictly ordered
in terms of termination properties: 
non-strict, function strict, and application strict.
% non-strict; function strict, in which only
%those arguments needed by a function call are evaluated; and application
%strict, in which all arguments of an application are evaluated.  
The other dimension we vary is whether or not constructors are strict.  Other
dimensions are possible, for example whether \texttt{let} bindings are strict,
but we restrict ourselves to function/application and constructor strictness.
The STG machinery makes the choice of evaluation strategy easy to change,
though changing code generation can be more efficient by avoiding the
unnecessary creation of \emph{THUNKS} as discussed in Appendix~A.

@@

\emph{EA} describes a family of generic C-{}- \emph{stgApply} runtime
functions used for applications of non-known (not statically identifiable)
functions, and applications of known functions to other than their arity
number of arguments.  There the object to be applied is first evaluated with
call-with-return.  When a resulting \emph{FUN} is applied to more than its
arity number of arguments, it is called-with-return with its arity number of
arguments, and the resulting value applied to the remaining arguments.  
%
Call-with-return becomes even more convenient for strict evaluation: arguments
can be evaluated in a tight loop, and similarly for strict constructors, if
the STG generator (human or otherwise) has not arranged for them to be already
evaluated.
%When considering more strict semantics call-with-return is even more
%convenient because arguments can be evaluated in a tight loop.

@@

Type information is required for code generation.  For example, at the
coarsest level it is necessary to know whether values are boxed, and so
represented by pointers, or unboxed, to avoid expensive tagging.  Because we
do not yet get type information directly from GHC we use a subset of Haskell
syntax (e.g., no named fields) for defining algebraic data types, extended
with built-in and user-defined unboxed types.  We perform standard
Hindley-Milner type inference and enforcement following Heren et
al.~\cite{Heren02}, again extended with unboxed types.  The extensions to the
type inference algorithm, syntax, and restrictions for unboxed types follow
Peyton Jones and Launchbury~\cite{Jones:1991}.
%\begin{comment}
% breaks the page limit
For convenience we allow general recursive \texttt{let}s that are implicitly
factored in the usual way to maximize let-polymorphism; this would normally
already be performed by a front-end.
%\end{comment}
\begin{comment}
 breaks the page limit
One would expect an automated STG generator such as GHC to factor recursive
\texttt{let} expressions into nested recursive and non-recursive \texttt{let}s
in the usual way to maximize let-polymorphism.  Because our STG is intended
to be somewhat convenient to write by hand we relax this constraint and perform
this transformation implicitly using a strongly-connected components algorithm
inspired by King and Launchbury~\cite{King:1995}.
\end{comment}

@@

Previously cited papers describing some of the inner workings of GHC at
various points in its evolution have greatly informed our approach.

\subsubsection{Strict but still pure.}
The concept of strict but still pure in some sense dates back to the concept
of applicative order reduction in the lambda calculus.  In terms of a
notionally concrete programming language the first thorough explications may
be Backus' FP~\cite{Backus:1978} and follow-on FL
implementation~\cite{AikenFL,FLreport89}.
%
A notable argument for the potential merits of a strict-by-default
pure functional language over a non-strict one is by Sheard, who also provided
a very simple prototype implementation~\cite{Sheard:2003}.
%
A plug-in for GHC (prior to 8.0) exists that makes function bodies strict in
their non-recursive \texttt{let} bindings using a transformation on GHC Core
that recursively expands those \texttt{let} bindings into \texttt{case}
bindings~\cite{Bolingbroke:2008,strict-ghc-plugin}.
%
The Disciplined Disciple Compiler project appears to intend implementation of
a strict, higher order, functional language (or family of languages) compiler
and has been underway for some time, but the emphasis appears to be on a much
more complex type system for handling stateful computation, and the prospects
for its eventual completion are unknown~\cite{disciplined-disciple}.

@@

Don Stewart described an implementation wherein GHC was used as a Haskell compiler
front-end, dumping STG code as text, parsing that text, then translating to
code for execution on the Java Virtual Machine~\cite{Stewart-BSc}.

@@

\section{Student Development}

Student interns' backgrounds in computer programming has ranged from minimal
to some exposure to one or more mainstream imperative languages, Lisp, and
Haskell.  To bring them to a normalized level of proficiency their first
assignment has been to work through the entirety of Hutton's \emph{Programming
  in Haskell} textbook~\cite{Hutton-book}, substituting one of his articles
\emph{Higher-order Functions for Parsing}~\cite{Hutton-parsing:1992} or
\emph{Monadic Parser Combinators}~\cite{Hutton:1996} for the corresponding
material in the book.  This may be followed by writing a parser and ADT for a
progressively enriched lambda calculus, together with call-by-value,
call-by-name, and (in one case) call-by-need interpreters.  The concept of
compilation can be illustrated with an exercise in parsing infix arithmetic
expressions to an AST, and transforming that to a sequence of operations for a
simple stack machine.
% breaks page limit
%  This learning phase can take the
%better part of a summer or one-semester internship, so we particularly
%seek return interns.

>>>>

\renewcommand{\floatpagefraction}{0.8}
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Automatic Parallelization and Transparent Fault Tolerance\\
       (\emph{New project} paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and
        Dean Prichard\inst{1} \and
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}%\thanks{Ringo, Anderson, and Marks are undergraduate students.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

A claimed trend is an increasing appreciation and practice of functional
programming in scientific computing, and in particular strict (by default)
functional programming, in various guises.  Besides making programs easier to
reason about, this enables automatic parallelization to various degrees.  A
second claimed trend in both scientific computing and functional programming
is transparent fault tolerance.  Our project goal is the demonstration of a
light-weight, higher-order, polymorphic, pure functional language
implementation in which we can experiment with automatic parallelization
strategies, varying degrees of strictness, and mechanisms for transparent
fault tolerance.
%
We do not consider speculative evaluation, or semantic strictness inferred by
program analysis, so potential parallelism is dictated by the specified degree of
strictness.

\hspace{0.2in} Our development has been greatly informed by the extant
literature and we indicate both where we follow, and deviate from,
these sources.
%
\\
{\color{OliveGreen} % \bf
\emph{Draft paper notes:}
\begin{enumerate}[before=\itshape]
\item We are reporting work in progress.  At the time of writing we have a
  serial implementation, a primitive proof-of-concept parallel
  implementation, and are currently developing a more realistic parallel
  runtime.  We expect to have preliminary results from our new parallel
  implementation by the time of the symposium.  The fault-tolerance mechanisms
  remain future work.

\item Hopefully in the spirit of a student-friendly symposium, this draft is
  liberally annotated with indicators of student contributions, student
  preparation, the influence of their involvement on our approach, and other
  incidental and ancillary comments.  A final paper would of course be cleaned
  up and potentially considerably shortened.

\item While \textcolor{OrangeRed}{this does not qualify as a student paper}, students
  (all undergraduate) have contributed considerably to this effort, and to the
  writing of this paper, and will continue to do so.  If this paper is
  accepted we plan to have one of these students present it at TiFP'16.
\end{enumerate}
}

\keywords{Strict pure functional programming, automatic parallelization,
  transparent fault tolerance, Haskell, STG, GHC, GHC Core.}

@@

\section{Background and Motivation}
The practice of so-called high-performance scientific computing (HPC) is,
overall, highly conservative with respect to change.  Even as the USA and
other governments commit to a push for exascale computing ($10^{18}$ FLOPS in
a single system) within the next
decade~\cite{Exascale-roadmap,EUexascale,Exascale-org} a respected
member of the HPC community recently stated publicly that ``Fortran is
essential for exascale programming''~\cite{Heroux16}.  Existing US government
laboratory HPC codes run to hundreds of thousands of lines of Fortran each and
complete rewrites are infeasible.\footnote{The codes of our lab's British
  counterpart are similar.}  C++ has become more common for new code starts,
but their nature remains the same: huge codes with fair to non-existent
encapsulation of side effects, with multiple levels of parallelism sometimes
poorly abstracted, especially at the thread (shared-memory) and computational
accelerator (e.g.\ GPU) levels.  Thread-level parallel efficiency can be good
when the parallel model is bulk synchronous; tasking models are generally
avoided unless abstracted by, e.g., a C++ template or runtime library, because
of the programming complexity.

@@

A common complaint among these aspiring high-performance functional
programmers is the need to get strictness just right: to remove space leaks,
achieve competitive performance, get I/O and inter-process communication to work
as they expect, and so on.  More specifically, they expect to be able to
reason about space and time usage, and order of evaluation, using their
existing mental models. While they are happy to write code that is strongly
reminiscent of the numerical examples in Hughes' classic piece (whether
they've read it or not)~\cite{Hughes89}, they would also be happy to write in
a slightly modified style that does not require non-strictness---after all,
that's how they've been thinking their entire careers.

@@

Our initial requirements are that
\begin{itemize}
\item The implementation must be feasible in terms of available effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling must be properly and consistently implemented;
\item The generated code is C.
\end{itemize}
%
For feasibility we restrict ourselves to implicit parallelism on shared-memory
multiprocessors.  The requirement for tail calling is not primarily for
efficiency, rather for an efficient parallel implementation.  Achieving
this, and thereby sufficiently limiting the use of the native C stack, will be
discussed at some length.

@@

\section{Design and Implementation}
Our wish list has much in common with the GHC Haskell implementation, and
indeed GHC is part of our master plan.  In brief, the phases of the GHC
compiler are (1) parsing and type checking; (2) transformation to the GHC
\emph{Core} intermediate representation~\cite{Sulzmann:2007,ghc-core}; (3)
analysis and transformation on Core; (4) transformation of Core to the
\emph{STG} intermediate representation~\cite{PJ-stockhardware}; and (5)
transformation of STG to machine code via various imperative
representations (Cmm, LLVM, C, native code generation).

@@

Greater strictness implies more opportunities for automatic parallelism,
possible losses or gains in space and time efficiency, and less expressiveness.  We
seek to explore the interplay of all of these in various ways.

GHC 8.0 gives three defaults: conventional non-strict, constructor strict, and
function- and constructor strict, but there are possible variations.  The
obvious variation is function, but not constructor, strictness.  In Haskell
the distinction between constructors and functions is somewhat blurred
(outside of data definitions and pattern matching) but in STG the distinction
is clear.

@@

In STG the arity of a function is defined as the number of formal parameters
in the function definition, thus \texttt{id x = x} has by definition arity one
regardless of the type of the argument it might be applied to (e.g., some
function type).  We distinguish \emph{application}, e.g., \texttt{f x y},
which is a syntactic notion, from actual function \emph{call}. Thus \texttt{id
  id x} is an application of \texttt{id} to two arguments, but the underlying
\emph{FUN} (user- or system- defined function) when called is given exactly
one argument (the first one).

In the eval-apply model, given application \texttt{f x y}, in the general case, with standard
(non-strict) semantics, \texttt{f} is first evaluated and the
arity of the underlying \emph{FUN} or \emph{PAP} (partial application of a
\emph{FUN}) is determined---in the case of \emph{PAP} the arity is that of the
underlying \emph{FUN} less the number of arguments previously provided.  If
the arity is greater than two a new \emph{PAP} is created with arguments
\texttt{x} and \texttt{y} inserted and the new \emph{PAP} returned.  If the
arity is equal to two the underlying \emph{FUN} is tail-called (jumped to)
with any previous arguments (in case of \emph{PAP}) and new arguments
\texttt{x} and \texttt{y}.  If the arity is less than two (it must be at least
one, so one in this case) the underlying \emph{FUN} is called with any
previous arguments (in case of \emph{PAP}) and new argument \texttt{x}, then
the object returned applied to remaining argument \texttt{y}.

Given application \texttt{f x y}, when might \texttt{x} and \texttt{y} be
evaluated?  In the non-strict semantics evaluation is on demand, after
\texttt{f} has been evaluated, the underlying \emph{FUN} actually called
(i.e., not just partially applied and wrapped up in a \emph{PAP}), and one or
both of those arguments possibly forced in subsequent evaluation, assuming
that they weren't forced in the evaluation of \texttt{f} itself.

In a more strict semantics one could devise various evaluation orders, but as far
as distinguishing termination properties it seems that these come
down to two reasonable possibilities, namely whether on construction of a
\emph{PAP} the (new) arguments are forced or not.  In other words, the
question is whether application is strict, or function call is strict.  These
are in fact distinguishable because (in Haskell, for example) the first
argument to \texttt{seq} could evaluate to a \emph{PAP}.

This then gives three reasonable operational modes which are strictly ordered
in terms of termination properties: 
non-strict, function strict, and application strict.
% non-strict; function strict, in which only
%those arguments needed by a function call are evaluated; and application
%strict, in which all arguments of an application are evaluated.  
The other dimension we vary is whether or not constructors are strict.  Other
dimensions are possible, for example whether \texttt{let} bindings are strict,
but we restrict ourselves to function/application and constructor strictness.
The STG machinery makes the choice of evaluation strategy easy to change,
though changing STG generation can be more efficient by avoiding the
unnecessary creation of \emph{THUNKS} as discussed in Appendix~A.

@@

\emph{EA} describes a family of generic C-{}- \emph{stgApply} runtime
functions used for applications of non-known (not statically identifiable)
functions, and applications of known functions to other than their arity
number of arguments.  There the object to be applied is first evaluated with
call-with-return.  When a resulting \emph{FUN} or \emph{PAP} is applied to
more than its arity number of arguments, it is called-with-return with its
arity number of arguments, and the resulting value applied to the remaining
arguments.
%
Call-with-return becomes even more convenient for strict evaluation: arguments
can be evaluated in a tight loop, and similarly for strict constructors, if
the STG generator (human or otherwise) has not arranged for them to be already
evaluated.
%When considering more strict semantics call-with-return is even more
%convenient because arguments can be evaluated in a tight loop.

@@

Type information is required for code generation.  For example, at the
coarsest level it is necessary to know whether values are boxed, or otherwise
represented by pointers, to avoid expensive tagging.  Because we
do not yet get type information directly from GHC we use a subset of Haskell
syntax (e.g., no named fields) for defining algebraic data types, extended
with built-in and user-defined unboxed types.  We perform standard
Hindley-Milner type inference and enforcement following Heren et
al.~\cite{Heren02}, again extended with unboxed types.  The extensions to the
type inference algorithm, syntax, and restrictions for unboxed types follow
Peyton Jones and Launchbury~\cite{Jones:1991}.
%\begin{comment}
% breaks the page limit
For convenience we allow general recursive \texttt{let}s that are implicitly
factored in the usual way to maximize let-polymorphism; this would normally
already be performed by a front-end.
%\end{comment}
\begin{comment}
 breaks the page limit
One would expect an automated STG generator such as GHC to factor recursive
\texttt{let} expressions into nested recursive and non-recursive \texttt{let}s
in the usual way to maximize let-polymorphism.  Because our STG is intended
to be somewhat convenient to write by hand we relax this constraint and perform
this transformation implicitly using a strongly-connected components algorithm
inspired by King and Launchbury~\cite{King:1995}.
\end{comment}

@@

\emph{Nihil sub sole novum}---there is nothing new under the sun (Ecclesistes 1:9).
%every new idea has some sort of precedent or echo from the past~\cite{nihil}.

%Previously cited papers describing some of the inner workings of GHC at
%various points in its evolution have greatly informed our approach.

\subsubsection{Strict but still pure.}
The concept of strict but still pure in some sense dates back to the concept
of applicative order reduction in the lambda calculus.  In terms of a
notionally concrete programming language the first thorough explications may
be Backus' FP~\cite{Backus:1978} and follow-on FL
implementation~\cite{AikenFL,FLreport89}.
%
A notable argument for the potential merits of a strict-by-default
pure functional language over a non-strict one is by Sheard, who also provided
a very simple embedded implementation~\cite{Sheard:2003}.
%
A plug-in for GHC (prior to 8.0) exists that makes function bodies strict in
their non-recursive \texttt{let} bindings using a transformation on GHC Core
that recursively expands those \texttt{let} bindings into \texttt{case}
bindings~\cite{Bolingbroke:2008,strict-ghc-plugin}.
%
The Disciplined Disciple Compiler project appears to intend implementation of
a strict, higher order, functional language (or family of languages) compiler
and has been underway for some time, but the emphasis appears to be on a much
more complex type system for handling stateful computation, and the prospects
for its eventual completion are unknown~\cite{disciplined-disciple}.

@@

Don Stewart described an implementation wherein GHC was used as a Haskell
compiler front-end, dumping STG code as text, parsing that text, then
translating to Java code for compilation and execution on the Java Virtual
Machine~\cite{Stewart-BSc}.

@@

\section{Student Preparation}

Student interns' backgrounds in computer programming has ranged from minimal
to some exposure to one or more mainstream imperative languages, Lisp, and
Haskell.  To bring them to a normalized level of proficiency their first
assignment has been to work through the entirety of Hutton's \emph{Programming
  in Haskell} textbook~\cite{Hutton-book}, substituting one of his articles
\emph{Higher-order Functions for Parsing}~\cite{Hutton-parsing:1992} or
\emph{Monadic Parser Combinators}~\cite{Hutton:1996} for the corresponding
material in the book.  This may be followed by writing a parser and ADT for a
progressively enriched lambda calculus, together with call-by-value,
call-by-name, and (in one case) call-by-need interpreters.  The concept of
compilation can be illustrated with an exercise in parsing infix arithmetic
expressions to an AST, and transforming that to a sequence of operations for a
simple stack machine.  From this point they have then been able to substantively
contribute to the project.

% breaks page limit
%  This learning phase can take the
%better part of a summer or one-semester internship, so we particularly
%seek return interns.
